---
title: "U.S. national happiness post-covid"
author: "Julie Nguyen, Missy Zhang, Linrong Li"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
format:
  pdf:
    documentclass: article
    papersize: a4
    geometry: margin = 1in
    mainfont: "Garamond"
    sansfont: "Garamond"
thanks: 'Code and data are available at: '
abstract: "add abstract"
number-sections: true
execute: 
  message: false
  warning: false
  echo: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(dplyr)
library(haven)
library(janitor)
library(tidyverse)
library(knitr)
library(vtable)
library(kableExtra)
library(readr)
library(ggplot2)
library(here)
library(readxl)
library(stringr)

raw_data <- haven::read_dta("../../inputs/data/gss2021.dta")
raw_data_job <- read_excel(here::here("inputs/job_happy/GSS.xlsx"))

# Clean and recode data
cleaned_data <- raw_data %>% 
  select(age, race, sex, happy) %>% 
  arrange(age) %>% 
  mutate(age = case_when(
    age < 18 ~ '< 18',
    age >= 18 & age <= 34 ~ '18-34',
    age >= 35 & age <=49 ~ '35-49',
    age >= 50 & age <=64 ~ '50-64',
    age >= 65 & age <=89 ~ '65 and above'
  )) %>% 
  mutate(sex = case_when(
    sex == 1 ~ 'Male',
    sex == 2 ~ 'Female'
  )) %>% 
  mutate(race = case_when(
    race == 1 ~ 'White',
    race == 2 ~ 'Black',
    race == 3 ~ 'Other'
  )) %>% 
  mutate(happy = case_when(
    happy == 1 ~ 'Very Happy',
    happy == 2 ~ 'Pretty Happy',
    happy == 3 ~ 'Not Too Happy'
  )) %>% 
  filter(!is.na(happy))

# Remove Inapplicable, Dont know, no answer, refused and skipped
cleaned_data_job <- raw_data_job %>% 
  filter(!str_detect(hrs1, '.i'), !str_detect(hrs1, '.n'), 
         !str_detect(hrs1, '.d'), !str_detect(hrs1, '.s'),
         !str_detect(rincome, '.i'), !str_detect(rincome, '.n'), 
         !str_detect(rincome, '.d'), !str_detect(rincome, '.s'),
         !str_detect(rincome, '.r'), !str_detect(happy, '.i'),
         !str_detect(happy, '.n'), !str_detect(happy, '.d'),
         !str_detect(happy, '.s'))
cleaned_data_job$rincome <- sub("-","TO",cleaned_data_job$rincome)
cleaned_data_job$rincome <- sub("LT","LOWER THEN ",cleaned_data_job$rincome)
cleaned_data_job$hrs1 <- as.numeric(cleaned_data_job$hrs1)
cleaned_data_job$rincome <- factor(cleaned_data_job$rincome,
                             levels = c('LOWER THEN  $1000',
                                        '$1000 TO 2999',
                                        '$3000 TO 3999',
                                        '$4000 TO 4999',
                                        '$5000 TO 5999',
                                        '$6000 TO 6999',
                                        '$7000 TO 7999',
                                        '$8000 TO 9999',
                                        '$10000 TO 14999',
                                        '$15000 TO 19999',
                                        '$20000 TO 24999',
                                        '$25000 OR MORE'))
```

## Introduction

## Data

### Source


Variables
```{r}
var <- raw_data %>% 
  select(age, sex)
var$sex <- var$sex %>%  to_factor()
sumtable(var, title = 'U.S. happiness in 2021', out = 'kable', simple.kable = TRUE)
```

### Methodology

### Ethical Concerns & Limitations

## Discussion

### National Happiness

In overall, the level of happiness in the United States in 2021 is "pretty happy". Among 4041 respondents, more than half (2308 respondents) reported that they feel happy, while 923 people were not happy and 783 people were extremely happy.

```{r}
cleaned_data %>% ggplot(aes(happy)) +
  geom_bar() +
  theme_minimal() +
  labs(x = "Level of Happiness",
       y = "Number of Respondents",
       title = "Overall national happiness")
```
<<<<<<< HEAD
  In order to understand the context and trends, we look into the demographics such as age, sex, and race.
  Comparing happiness among different age groups, it is apparent that the level of happiness remained stable throughout the groups. While in 18 to 64 or non-identifiable age groups, there are more people feeling unhappy than extremely happy, from 65 and above age group witnesses the same proportion of people feeling not too happy and extremely happy. 
  
=======

We also look into the demographics such as age, sex, and race. Comparing happiness among different age groups, it is apparent that the level of happiness remained stable throughout the groups. While in 18 to 64 or non-identifiable age groups, there are more people feeling unhappy than extremely happy, from 65 and above age group witnesses the same proportion of people feeling not too happy and extremely happy.

>>>>>>> 49e7b61acae834bc726ffcda98ca006ea8ce7d66
```{r}
# Happiness by age group
cleaned_data %>% group_by(age) %>% 
  ggplot(aes(x=happy, fill=age)) +
  geom_bar() +
  facet_wrap(~age, scales ="free") +
  theme_minimal() +
  labs(x="Level of Happiness",
       y="Number of Respondents",
       title="Happiness by age")+
  theme(legend.position = "none")
```

Looking at different sex groups, male tend to be more happy than female.

```{r}
# Happiness by sex
cleaned_data %>% group_by(sex) %>% 
  ggplot(aes(x=happy, fill=sex)) +
  geom_bar() +
  facet_wrap(~sex, scales ="free") +
  theme_minimal() +
  labs(x="Level of Happiness",
       y="Number of Respondents",
       title="Happiness by sex")+
  theme(legend.position = "none")
```

```{r}
# Happiness by race
cleaned_data %>% group_by(race) %>% 
  ggplot(aes(x=happy, fill=race)) +
  geom_bar() +
  facet_wrap(~race, scales ="free") +
  theme_minimal() +
  labs(x="Level of Happiness",
       y="Number of Respondents",
       title="Happiness by race")+
  theme(legend.position = "none")
```

Besides we looked into how job conditions may affect the level of happiness.

Firstly, we looked at the hours the respondent worked in the last week from they responded to the survey.

```{r}
#| echo: false
#| label: tbl-hour
#| tbl-cap: The means of hours worked last week in different levels of happiness
hrs_happy <- cleaned_data_job %>% 
  group_by(happy) %>% 
  summarize(Mean = mean(hrs1),
            Sd = sd(hrs1),
            Total_observed = n()) %>% ungroup()

knitr::kable(hrs_happy)
```

Then, we looked at how people in different income group rated their levels of happiness.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| label: tbl-income
#| tbl-cap: The level of happiness in different income groups
income_happy <- cleaned_data_job %>%
  group_by(rincome) %>%
  count(happy) %>%
  ungroup()

income_happy_table <- kbl(income_happy) %>%
  collapse_rows(columns = 1:2, valign = "top")
income_happy_table
```

To better visualize the results and find the tendency, we plotted a graph.

```{r}
#| echo: false
#| fig-width: 20
#| fig-height: 15
#| label: fig-income
#| fig-cap: Relationship of income level and happiness
cleaned_data_job  %>%  group_by(rincome) %>% 
  ggplot(aes(x = happy, fill = rincome)) + 
  geom_bar() +
  facet_wrap(~rincome, scales = "free") +
  theme_minimal() + 
  labs(x = "Level of Happiness of Respondent", 
       y = "Number of Respondents",
       title = "Relationship of Income Respondent and Happiness") + 
  theme(legend.position = "none")
```

## Conclusion

## Reference

## Appendix
