---
title: "U.S. national happiness post-covid"
author: "Julie Nguyen, Missy Zhang, Linrong Li"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
format:
  pdf:
    documentclass: article
    papersize: a4
    geometry: margin = 1in
    mainfont: "Garamond"
    sansfont: "Garamond"
thanks: 'Code and data are available at: '
abstract: "add abstract"
number-sections: true
execute: 
  message: false
  warning: false
  echo: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(dplyr)
library(haven)
library(janitor)
library(tidyverse)
library(knitr)
library(vtable)
library(kableExtra)
library(readr)
library(ggplot2)
library(here)
library(readxl)
library(stringr)
library(labelled)

raw_data <- haven::read_dta("../../inputs/data/gss2021.dta")
raw_data_job <- read_excel(here::here("inputs/job_happy/GSS.xlsx"))

# Clean and recode data
cleaned_data <- raw_data %>% 
  select(age, race, sex, happy) %>% 
  arrange(age) %>% 
  mutate(age = case_when(
    age < 18 ~ '< 18',
    age >= 18 & age <= 34 ~ '18-34',
    age >= 35 & age <=49 ~ '35-49',
    age >= 50 & age <=64 ~ '50-64',
    age >= 65 & age <=89 ~ '65 and above'
  )) %>% 
  mutate(sex = case_when(
    sex == 1 ~ 'Male',
    sex == 2 ~ 'Female'
  )) %>% 
  mutate(race = case_when(
    race == 1 ~ 'White',
    race == 2 ~ 'Black',
    race == 3 ~ 'Other'
  )) %>% 
  filter(!is.na(happy))

# Remove Inapplicable, Dont know, no answer, refused and skipped
cleaned_data_job <- raw_data_job %>% 
  filter(!str_detect(hrs1, '.i'), !str_detect(hrs1, '.n'), 
         !str_detect(hrs1, '.d'), !str_detect(hrs1, '.s'),
         !str_detect(rincome, '.i'), !str_detect(rincome, '.n'), 
         !str_detect(rincome, '.d'), !str_detect(rincome, '.s'),
         !str_detect(rincome, '.r'), !str_detect(happy, '.i'),
         !str_detect(happy, '.n'), !str_detect(happy, '.d'),
         !str_detect(happy, '.s'))
cleaned_data_job$rincome <- sub("-","TO",cleaned_data_job$rincome)
cleaned_data_job$rincome <- sub("LT","LOWER THEN ",cleaned_data_job$rincome)
cleaned_data_job$hrs1 <- as.numeric(cleaned_data_job$hrs1)
cleaned_data_job$rincome <- factor(cleaned_data_job$rincome,
                             levels = c('LOWER THEN  $1000',
                                        '$1000 TO 2999',
                                        '$3000 TO 3999',
                                        '$4000 TO 4999',
                                        '$5000 TO 5999',
                                        '$6000 TO 6999',
                                        '$7000 TO 7999',
                                        '$8000 TO 9999',
                                        '$10000 TO 14999',
                                        '$15000 TO 19999',
                                        '$20000 TO 24999',
                                        '$25000 OR MORE'))
```

## Introduction

## Data

### Source


Variables
```{r}
var <- raw_data %>% 
  select(age, sex)
var$sex <- var$sex %>%  to_factor()
sumtable(var, title = 'U.S. happiness in 2021', out = 'kable', simple.kable = TRUE)
```

### Methodology

### Ethical Concerns & Limitations

## Discussion
  In order to understand the context and trends, we observed the overall trend of happiness of the U.S. population in 2021. We also looked into the distribution of happiness among different age groups, sex groups, and racial groups.

### National Happiness

  In overall, the level of happiness in the United States in 2021 is "pretty happy". Among 4041 respondents, more than half (57.5%) reported that they feel happy, while 923 people were not happy and 783 people were extremely happy. [@fig-overallhp]

```{r, fig.pos="H"}
#| label: fig-overallhp
#| fig-cap: "National Happiness"
overall_hp <- cleaned_data %>% 
  group_by(happy) %>% 
  count(happy)
overall_hp <- overall_hp %>% mutate(proportion = n/sum(n)*100)

cleaned_data %>% ggplot(aes(happy)) +
  geom_bar() +
  theme_minimal() +
  labs(x = "Level of Happiness",
       y = "Number of Respondents")
```
### Happiness among age groups
  Comparing happiness among different age groups, it is apparent that the level of happiness remained stable throughout the groups. While in 18 to 64 or non-identifiable age groups, there are more people feeling unhappy than extremely happy, from 65 and above age group witnesses the same proportion of people feeling not too happy and extremely happy.

```{r, fig.pos="H"}
# Happiness by age group
cleaned_data %>% group_by(age) %>% 
  ggplot(aes(x=happy, fill=age)) +
  geom_bar() +
  facet_wrap(~age, scales ="free") +
  theme_minimal() +
  labs(x="Level of Happiness",
       y="Number of Respondents",
       title="Happiness by age")+
  theme(legend.position = "none")
```

### Happiness among male, female, and others 

@fig-bysex shows that male tend to be more happy than female. In the year 2021, the U.S. population reported to relatively happy among different sex groups. There is a great disparity between female respondents feeling unhappy and very happy - more female experienced unhappiness than extremely happy, while that of male respondents is quite similar. In addition, there are 92 people who responded their level of happiness without identifying their sex. In this group, there are 62 respondents feel happy, 18 respondents feel extremely happy, and 12 respondents feel unhappy.  

```{r, fig.pos="H"}
#| label: fig-bysex
#| fig-cap: "Happiness by sex in 2021"

# Happiness by sex
cleaned_data %>% group_by(sex) %>% 
  ggplot(aes(x=happy, fill=sex)) +
  geom_bar() +
  facet_wrap(~sex, scales ="free") +
  theme_minimal() +
  labs(x="Level of Happiness",
       y="Number of Respondents",
       title="Happiness by sex")+
  theme(legend.position = "none")
```
### Happiness among different races
  As a multicultural country, happiness in different races is also worth observing as it shows the level of happiness in different racial communities. From @fig-race We notice that there is a similar pattern between the black and white racial groups, and between other races and non-identifiable group.

```{r}
#| label: fig-race
#| fig-cap: "Happiness by race"

# Happiness by race
cleaned_data %>% group_by(race) %>% 
  ggplot(aes(x=happy, fill=race)) +
  geom_bar() +
  facet_wrap(~race, scales ="free") +
  theme_minimal() +
  labs(x="Level of Happiness",
       y="Number of Respondents")+
  theme(legend.position = "none")
```
### Happiness and job conditions
Besides we looked into how job conditions may affect the level of happiness.

Firstly, we looked at the hours the respondent worked in the last week from they responded to the survey.

```{r}
#| echo: false
#| label: tbl-hour
#| tbl-cap: The means of hours worked last week in different levels of happiness
hrs_happy <- cleaned_data_job %>% 
  group_by(happy) %>% 
  summarize(Mean = mean(hrs1),
            Sd = sd(hrs1),
            Total_observed = n()) %>% ungroup()

knitr::kable(hrs_happy)
```

Then, we looked at how people in different income group rated their levels of happiness.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| label: tbl-income
#| tbl-cap: The level of happiness in different income groups
income_happy <- cleaned_data_job %>%
  group_by(rincome) %>%
  count(happy) %>%
  ungroup()

income_happy_table <- kbl(income_happy) %>%
  collapse_rows(columns = 1:2, valign = "top")
income_happy_table
```

To better visualize the results and find the tendency, we plotted a graph.

```{r}
#| echo: false
#| fig-width: 20
#| fig-height: 15
#| label: fig-income
#| fig-cap: Relationship of income level and happiness
cleaned_data_job  %>%  group_by(rincome) %>% 
  ggplot(aes(x = happy, fill = rincome)) + 
  geom_bar() +
  facet_wrap(~rincome, scales = "free") +
  theme_minimal() + 
  labs(x = "Level of Happiness of Respondent", 
       y = "Number of Respondents",
       title = "Relationship of Income Respondent and Happiness") + 
  theme(legend.position = "none")
```

## Conclusion

## Reference

## Appendix
